---
title: "Research Computing Bootcamp"
author: "Boston University"
output:
  html_document:
    df_print: paged
---

<style type="text/css">

body{ /* Normal  */
  font-size: 12px;
}
td {  /* Table  */
  font-size: 10px;
}
code.r{ /* Code block */
    font-size: 12px;
    background-color: #FAEAEA; 
}
pre { /* Code block - determines code spacing between lines */
    font-size: 12px;
}

.more {
  color: #580000;
  font-style: italic;
}
</style>

<br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source="bg-success")
```

## R Track: Part 3 - Data Visualization in R


Examining complex patterns and correlations in data requires visualization to present the results in a digestible form. 
Part Three presents various ways to plot and visualize data. 
Different plotting techniques and formats are presented, along with various ways to format and configure their display and appearance.

R plotting libraries are designed to produce high-quality graphics.

Various types of plots allow to generate plots of:

- a single variable
- two variables
- multiple variables
- networks
- time series
- many other special plots


First, we need to load the packages that we will use for this workshop"
```{r, warning=FALSE, message=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
```

<br/><br/>

Let's read the two dataset we explored in the previous workshop and merg them together:


```{r, warning=FALSE, message=FALSE}
NE.health <- read_csv("https://github.com/bu-rcs/bu-rcs.github.io/raw/main/Bootcamp/Data/NE_HealthData.csv") %>%
  rename(poor_health = "% Fair or Poor Health",
         p_smokers = "% Smokers",
         p_obese = "% Adults with Obesity",
         p_inactive = "% Physically Inactive",
         p_college="% Some College",
         p_housing_problems = "% Severe Housing Problems") %>%
  select(State, County, poor_health : p_inactive, p_college, p_housing_problems) %>%
  filter( !is.na(County) )
  
  
  
NE.demogr <- read_csv("https://github.com/bu-rcs/bu-rcs.github.io/raw/main/Bootcamp/Data/NE_DemographicsData.csv") %>%
  rename( life_exp = "Life Expectancy", 
          ph_distress = "% Frequent Physical Distress",
          limit_healthy_food = "% Limited Access to Healthy Foods",
          house_income = "Median Household Income",
          N_black = "# Black",
          N_rural = "# Rural") %>%
  select(State : house_income, N_black, N_rural) %>%
  filter( !is.na(County) )


NE.data <- inner_join( NE.health, NE.demogr, by = c("State", "County"))
glimpse (NE.data)    # similar to str() but slightly different format of output


```

<br/><br/>

## Using R Graphics for data exploration


<br/><br/>

<h4 class="more">Read more:</h4>

- [ggplot2 tutorial](http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html)
- [ggplot2 cheat sheet ](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)
- [R graph gallery](https://www.r-graph-gallery.com/)
- [ggplot2 extensions](https://exts.ggplot2.tidyverse.org/gallery/)



<br/><br/>

### Homework 


<br/><br/>
