---
title: "Data Wrangling and Management"
author: "Research Computing Services"
output:
  html_document:
    df_print: paged
---

<style type="text/css">

body{ /* Normal  */
  font-size: 12px;
}
td {  /* Table  */
  font-size: 10px;
}
code.r{ /* Code block */
    font-size: 12px;
    background-color: #FAEAEA; 
}
pre { /* Code block - determines code spacing between lines */
    font-size: 12px;
}

.more {
  color: #580000;
  font-style: italic;
}
</style>

<br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source="bg-success")
```


To work with a dataset various operations are needed to bring it to a usable state within a program. During this workshop we will learn how to read data from a file, clean, format and manipulate the data. By the end of it you will learn how to:

We will be use *tibbles* which are similar to R's *data frames*.
*Tibbles* make R data manipulations much easier. They are defined in the **tibble** package
which is a part of the *tidyverse* set of packages.
To access the functions defined in an R package, we use a `library()` function:
```{r}
library(tidyverse)
```



<br/><br/>

### Reading Data using *readr* and *readxl* packages

*readr* package comes with its own set of functions for reading rectangular data (like 'csv', 'tsv', and fwf'). They are designed to parse many types of data. These functions generally 
faster than regular `read.csv()` and similar functions. To be able to use this functions, 
you can load *tidyverse* package or just *readr* package: `library(readr)`.

*readxl* packages allows reading excel-formatted files



```{r readdata }
# Read facebook pseudo data
library(readxl)
fb <- read_excel("pseudo_facebook.xlsx", sheet=1)
```

<br/><br/>

We can use `head()`, `str()`, and `summary()` functions to explore the dataset.
```{r}
# display class of the fb object
class(fb)
```

```{r}
# Explore the structure of fb object
str(fb)
```

<br><br>
*dplyr* package (part of tidyverse) has a similar function - `glimpse()`

```{r}
# Explore the structure of fb
glimpse(fb)
```

```{r}
# Summary for columns
summary(fb)
```

```{r}
# Find unique values of "gender" column
unique(fb$gender)
```
<br/><br/>

We need to make sure we handle missing values correctly. Let's re-read the dataset and
specify how missing values are coded in this dataset

```{r }
fb <- read_excel("pseudo_facebook.xlsx", sheet=1, na= c("","NA", "#N/A"))
```

<br/><br/>
Let's check unique values of the *gender* column again
```{r}
unique(fb$gender)
```

`table()` is a useful function to use for categorical variables:

```{r}
table(fb$gender, useNA="ifany")
```


<br/><br/>
Since `fb` is not only a *data.frame*, but also a *tibble*, some R functions
behave differently:

```{r}
print( fb )
```

<br/><br/>
When printing a tibble in a console window,  `print()` function  outputs only those observations and columns
that fit the Console window screen. We can modify this behavior using optional arguments:

```{r}

# print 12 observations and all columns
print( fb, n = 12, width = Inf )
```


Tidyverse packages use the following abbreviations to describe the type of each variable:

- *int* - integers
- *dbl* - real numbers (numeric or double precision)
- *chr* - character vectors (or strings)
- *dttm* - date and time vectors
- *date* - dates
- *lgl* - logical vectors
- *fct* - factors



<br/><br/>

### Data Transformations using *dplyr* package

Before we can analyze data, we often need to transform the data to get it in the right format.

To do this we might need to rename columns, filter some rows and possibly exclude some columns.
We might need to create some new variables or reorder the observations or columns.

All these operations can easily be done with a *dplyr* package (which is part of *tidyverse*)

There are five key dplyr functions:

- `filter()` - subset the datasets
- `select()` - select specific columns
- `mutate()` - modify existing or create new variables/columns
- `arrange()` - reorder rows
- `summarise()` - collapse dataset to a summary

Above functions can be used in conjunction with `group_by()` which allows perform operations
on the dataset split into groups.

In addition there are many other useful functions including:

- `rename()` - rename columns
- `count()` - count observations by group
- `case_when()` - vectorised *if()* function


<br/><br/>

### *rename()* - rename columns

We often need to simplify the names of columns to make them easier to use in the code.
`read_csv()` and other tidyverse functions that read data do not modify original column names 
and as a results those columns where special characters (like spaces or symbols) are used can
only be used within back ticks.

```{r}
# rename columns
fb1 <- rename(fb, mlikes = "mobile_likes", mlikes_received = "mobile_likes_received")
names( fb1 )
```


<br/><br/>

### *select()* - select columns of interest

```{r}
# select four columns 
fb2 <- select ( fb1, userid, age, gender:friendships_initiated )
glimpse(fb2)
```


<br/><br/>

### *filter()* - subsetting rows using column values



```{r}
# select observations where friend_count is greater than zero
fb3 <- filter ( fb2, friend_count > 0 )
head(fb3)
```

<br/>
Sometimes we need to further filter our dataset:

```{r}
# select only observations with gender column equal to "male"
fb3_male <- filter ( fb3, gender == "male" )
head(fb3_male)
```


Useful filter functions

- `==` (equal) and `!=` (not equal)
- `>`, `>=`, `<`, and `<=`
- is.na()
- between()

We can connect two or more logical constructs using

- `&` (logical *and*)
- `|` (logical *or*)
- `!` (negation)

<br/>

```{r}
# select observations for which age column does not have missing values and age 
# is less than 18 and gender is equal to "female"
fb3_young_females <- filter ( fb3, !is.na(age) & age<18 &  gender == "female")
summary(fb3_young_females)
```

<br/><br/>

### *mutate()* - create new or modify existing variables (columns)

```{r}
# Create a new column *adult* which will be equal to 1 if a person 18 years old or older
fb4 <- mutate ( fb, adult = ifelse(age < 18, 0, 1) )
head(fb4)
```

<br/><br/>

### *arrange()* - sort the dataset by values in one or more columns

```{r}
# Sort by age
fb5 <- arrange ( fb, age )
head(fb5)
```

```{r}
# Sort by age in reverse order
fb6 <- arrange ( fb, desc(age) )
head(fb6)
```


<br/><br/>

### *%>%* - Pipe symbol

In most cases we need to perform several operations on the input dataset before 
we can analyse it. For example we might need to use some or all functions, that
we learned so far - `rename()`, `filter()`, `mutate()`, `select()`, `arange()`.

So far we were saving each intermediate step in a separate variable (*fb2*, *fb3*, etc.)
This is not ideal as our input dataset might be large and each variable requires a separate space in memory.
It is also make it hard to remember which variable holds which data.frame.

*Tidyverse* packages use `%>%` (*pipe*) symbol to pass the result from one step to another.
Pipe symbol allows to take the dataset on the left side and pass it as the first argument to 
the function on the right side.

The following two ways of selecting three columns are equivalent

```{r, eval = FALSE}
select ( fb, userid, age, gender:likes)

fb  %>%  select ( userid, age, gender:likes)
```


Imagine that we need to perform the following operations after we read our data:

1. Rename columns
2. Select some columns
3. Filter rows (leaving only observations with 1 or more friend or friendship initiated)
4. Sort observations by "age" column in reverse order

We can use `%>%` symbol to do it in one operation:

<br/>

```{r}

fb_friends <- fb  %>% 
  rename ( friend_ini = "friendships_initiated") %>%
  select ( userid, age, gender: likes_received) %>%
  filter ( ! is.na(age) & (friend_count > 0 | friend_ini > 0 )) %>% 
  arrange ( desc(age) )
  
  head(fb_friends)
  
```

<br/><br/>

### *summarise()* - Calculate summary 

<br/>
```{r}
# Print number of observations  and
# mean value of friends

fb %>% summarise ( N = n(), 
                   ave_friends = mean(friend_count))
```


<br/><br/>


### *group_by()* - Grouping operator


*group_by()* function allows splitting datasets into groups, so we could apply other functions
like mutate, arrange, and summarise per group.

<br/>
```{r}
# Find average number of friends for male and female

fb  %>% 
  filter(!is.na(gender)) %>%
  group_by (gender ) %>% 
  summarise ( N= n(), ave_friends = mean (friend_count), .groups = "drop" )

```

<br/><br/>

Here is another example: We will create an "age group" variable and then use summarise
to find an average number of friends within each age group

```{r}

fb  %>% 
  filter(!is.na(age)) %>%
  mutate(age_group = case_when( between(age, 1, 17) ~ "child",
                                between(age, 18, 29) ~ "young adult",
                                between(age, 30, 69) ~ "adult",
                                TRUE ~ "senior")) %>%
  group_by (age_group ) %>% 
  summarise ( N= n(), ave_friends = mean (friend_count), .groups = "drop" )

```

<br/><br/>

<h4 class="more">Read more:</h4>

- [readr package chearsheet](https://github.com/rstudio/cheatsheets/blob/master/data-import.pdf)
- [Useful tips on using read_csv and other functions from readr package ](https://r4ds.had.co.nz/data-import.html)
- [Data Transformation using dplyr package](https://r4ds.had.co.nz/transform.html)
- [Merging two tables](https://r4ds.had.co.nz/relational-data.html)



<br/><br/>
