---
title: "Research Computing Bootcamp"
author: "Boston University"
output:
  html_document:
    df_print: paged
---

<style type="text/css">

body{ /* Normal  */
  font-size: 12px;
}
td {  /* Table  */
  font-size: 10px;
}
code.r{ /* Code block */
    font-size: 12px;
    background-color: #FAEAEA; 
}
pre { /* Code block - determines code spacing between lines */
    font-size: 12px;
}

.more {
  color: #580000;
  font-style: italic;
}
</style>

<br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source="bg-success")
```

## R Track: Part 4 - Using Statistical Tools for Data Analysis


Discovering and verifying meaningful patterns and correlations in data requires quantitative techniques that can analyze the data in a rigorous mathematical or algorithmic way. Part Four presents how to use various basic statistical tools to analyze the example dataset. The use of these tools is demonstrated and various example diagnostics, patterns, and correlations are examined in the example dataset.

```{r, warning=FALSE, message=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
```

<br/><br/>

Let's read the two dataset we explored in the previous workshops:


```{r, warning=FALSE, message=FALSE}
USA.health <- read_csv("https://github.com/bu-rcs/bu-rcs.github.io/raw/main/Bootcamp/Data/USA_HealthData.csv") %>%
  rename(poor_health = "% Fair or Poor Health",
         p_smokers = "% Smokers",
         p_obese = "% Adults with Obesity",
         p_inactive = "% Physically Inactive",
         p_college="% Some College",
         p_housing_problems = "% Severe Housing Problems") %>%
  select(State, County, poor_health : p_inactive, p_college, p_housing_problems) %>%
  filter( !is.na(County) )
  
  
  
USA.demogr <- read_csv("https://github.com/bu-rcs/bu-rcs.github.io/raw/main/Bootcamp/Data/USA_DemographicsData.csv") %>%
  rename( life_exp = "Life Expectancy", 
          ph_distress = "% Frequent Physical Distress",
          limit_healthy_food = "% Limited Access to Healthy Foods",
          house_income = "Median Household Income",
          N_black = "# Black",
          N_rural = "# Rural") %>%
  select(State : house_income, Population, N_black, N_rural) %>%
  filter( !is.na(County) )


USA.data <- inner_join( USA.health, USA.demogr, by = c("State", "County"))


USA.regions <- read_csv("https://github.com/bu-rcs/bu-rcs.github.io/raw/main/Bootcamp/Data/us_states_census_regions.csv")

# Add the names of Regions
USA.data <- merge(USA.data, USA.regions, by = "State")

glimpse (USA.data)   

```

<br/><br/>

### Hypothesis Testing in R

1. State Hypothesis
2. Formulate an analysis plan
3. Perform analysis
4. Interpret results


**Type I Error** - null hypothesis is rejected when it is true. *significance level* - probability of Type I error (represented by $\alpha$ symbol)

**Type II Error** - accepting a false null hypothesis. *power* of the test - probability of Type II error (represented by $\beta$ symbol)



<br/><br/>

### Checking if a variable normally distributed

Let's take a look at the *life expectancy* variable.

```{r, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}

ggplot( USA.data, aes (life_exp) ) + 
  geom_histogram(binwidth=1) + 
  geom_vline(xintercept=median(USA.data$life_exp, na.rm=T), 
             linetype="dashed", color = "red", size=1.) + 
  ggtitle ("USA Life expectancy distribution") + 
  xlab( "Life expectancy, years") +
  annotate(geom="text", 
           x=82, y=400, 
           label=round(median(USA.data$life_exp, na.rm=T),1),
           color="red") + 
  theme_bw()
  

```

<br/>
Is this variable normally distributed?

Let's first perform some *visual* analysis

```{r, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}

ggplot( USA.data, aes (life_exp) ) + 
  geom_density(color="darkgray", fill="lightgrey") + 
  geom_vline(xintercept=median(USA.data$life_exp, na.rm=T), 
             linetype="dashed", color = "red", size=.8) + 
  ggtitle ("USA Life expectancy distribution") + 
  xlab( "Life expectancy, years") +
  annotate(geom="text", 
           x=82, y=0.13, 
           label=round(median(USA.data$life_exp, na.rm=T),1),
           color="red") + 
  theme_bw()
  

```

<br/><br/>

```{r, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}

ggplot( USA.data, aes (sample = life_exp) ) + 
  stat_qq() + stat_qq_line() + 
  ggtitle ("qq-plot of  Life expectancy variable") + 
  theme_bw()
  

```

<br/>



1. **State Hypothesis**: 

*Null-hypothesis* - the sample comes from a normal distribution. *Alternative hypothesis* - the sample does not come from a normal distribution


2. **Formulate an analysis plan**

We will run [*Shapiro-Wilks*](http://www.sthda.com/english/wiki/normality-test-in-r) test of normality.


3. **Perform analysis**

```{r}
shapiro.test(USA.data$life_exp)
```

<br/>

4. **Interpret results**

From the output, the p-value < 0.05 implying that the distribution of the data 
is significantly different from normal distribution. 
So we reject the *Null Hypothesis* - data are not normally distributed



<br/><br/>

### Comparing two samples

Let's compare Life expectancy in New England and East South Ctr Region. 
Let's first summarize our data and find median and mean values

```{r}
USA.data %>%
  group_by(Region) %>%
  summarise ( mean = mean(life_exp, na.rm=T), median = median(life_exp, na.rm=T))

```

Let's visually compare two distributions of this variable for both regions


```{r, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}

ggplot( USA.data, aes (x=life_exp) ) + 
  geom_histogram(data = subset(USA.data, Region == "East South Ctr"), 
                 binwidth=1, fill = "red", alpha = 0.2) + 
  geom_histogram(data = subset(USA.data, Region == "New. Eng"), 
                 binwidth=1, fill = "blue", alpha = 0.2) + 
  geom_vline(xintercept=74.5, 
             linetype="dashed", color = "red", size=1.) + 
  geom_vline(xintercept=79.5, 
             linetype="dashed", color = "blue", size=1.) + 
  ggtitle ("USA Life expectancy in New England and East South Ctr Region") + 
  xlab( "Life expectancy, years") +
  annotate(geom="text", 
           x=75, y=77, 
           label=74.5,
           color="red") + 
    annotate(geom="text", 
           x=80, y=77, 
           label=79.5,
           color="blue") + 
  theme_bw()
  

```

<br/><br/>


1. **State Hypothesis**: 

*Null-hypothesis* - difference in means is equal to 0. *Alternative hypothesis* - true difference in means is not equal to 0.


2. **Formulate an analysis plan**

We will run [*Welch Two Sample t-test*](https://www.datanovia.com/en/lessons/types-of-t-test/unpaired-t-test/welch-t-test/).


3. **Perform analysis**

```{r}
x <- USA.data$life_exp[USA.data$Region == "New. Eng"]
y <- USA.data$life_exp[USA.data$Region == "East South Ctr"]
t.test(x,y)
```

<br/>

4. **Interpret results**

From the output, the p-value < 0.05 implying that the true difference in means is not
equal to zero and we can reject *Null Hypothesis*. It's always a good idea to include 
in your report the
95% percent confidence interval and other  statistics along with the p-value.


<br/><br/>

#### **Exercise**

1. Check if the variable *house_income* normally distributed
2. Perform Welch t.test to compare *house_income* between New England Region and East South Ctr



<br/><br/>


### Correlation and Covariance

From our previous sessions we saw that there is correlation between percent of smokers in a County
and *poor health* variable:


```{r, fig.width=6, fig.height=4}
ggplot( USA.data, aes(x = p_smokers, y = poor_health)) + 
  geom_point(alpha=0.1) + 
  geom_smooth(method = "lm") + 
    labs( title = "Adults with Poor Health vs. Percent of Smokers",
        subtitle="USA Population", 
        x =  "Smokers (%)",  
        y = "Poor Health (%)")
```

<br/>

The `cor()` function calculates correlations between two vectors. 

The `cov()` functions computes covariance. 

The `cor.test()` function performs test of significance of the correlation.


```{r}
cor( USA.data$p_smokers, USA.data$poor_health)

cov( USA.data$p_smokers, USA.data$poor_health)

cor.test( USA.data$p_smokers, USA.data$poor_health)

```

<br/><br/>

#### **Exercise**

1. Find correlation, covariance and perform test for correlation between *p_college* and 
*p_smokers* variables.



<br/><br/>

<h4 class="more">Read more:</h4>

- [T.test](http://www.sthda.com/english/wiki/t-test-analysis-is-it-always-correct-to-compare-means#t-test)
- [Correlation analysis](http://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r#correlation-formula)




<br/><br/>

### Homework 

1. Using violin plot explore the distribution of p_smokers variable per Region.
2. Add geom_jitter to better see the distribution of values for each region
3. Plot *life_exp* variable vs. *p_smokers*. Add title and modify the axes labels to make the graph more clear
4. Which State has a county with the life_exp value greater than 100? Explore the demographics in this region.


<br/><br/>


