---
title: "Iterate your code efficiently: Loops and Apply functions in R"
author: "Research Computing Services"
output:
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source="bg-success")
```

A few reasons to use loops in your code instead of copying-and-pasting:
1. reduce duplication
2. It's easier to read the code 
3. If part of the code needs to be changed, you only need to make changes in one place.
4. Code will have fewer bugs.

### Iterate over a range of integer values

```{r}

for (i in 1:3) {
  print(i)
}

```

  

### Iterate over a character vector

```{r}

for (i in month.name) {
  print(i)
}

```

### Using loops to avoid code duplication
  
  
**Task**: From Election file for each sentiment subset the records where name of Pres. Biden appears in FullText and save it
as a separate file
  
```{r, warning=FALSE, message=FALSE}
library(tidyverse)

# read the data
elect <- read_csv("election.csv") %>%
  select(Title, Sentiment, `Full Text`, Gender, Hashtags, Impact) %>%
  rename ( FullText = "Full Text") 


#check unique values in the column Sentiment
unique(elect$Sentiment)
```
  
  
  
```{r}

for ( sent in unique(elect$Sentiment) ) {
  
  sub <- elect %>% 
    filter(Sentiment == sent) %>%
    filter ( str_detect(FullText, fixed("Biden", ignore_case = T) ))
  print ( paste("Number of records in the subset for sentiment", sent, ":", nrow(sub) ))
  
}

```
  
## `lapply`
  
  
We often need to work with lists. For example, we can use function `strsplit()` to 
separate various hashtags for each observations. The result will be a list:

```{r}
hashtag_list <- strsplit(elect$Hashtags, split = ",")

``` 

We can now use `lapply` function to operate on each element of the list.
For example we can find the number of tags for each record:

```{r}
N_tags_list <- lapply( hashtag_list, length )
```

## `sapply`

List are sometimes hard to work with. And when it is possible we want to convert
them to vectors. In those cases we can use `sapply` function instead (`sapply()` is "simplified" version of `lapply()):

```{r}
N_tags <- sapply( hashtag_list, length )
```


## `tapply`

`tapply` function is used to apply a specific function to a column separating the dataset for each subgroup of another column.

For example if we want to compute the median number of tags for each sentiment, 
we could do:

```{r}
tapply( N_tags, elect$Sentiment, median )

```

Note, that in this case we could also use `dplyr` functions: `group_by()` and `summarise()`:

```{r}
elect %>% 
  mutate (N_tags = sapply(strsplit(Hashtags, ","), length)) %>%
  group_by(Sentiment) %>%
  summarise (median = median(N_tags) , .groups="drop")

```


