---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research Computing Bootcamp
### R Track: Part 1

This Bootcamp will teach you how to analyze a dataset using R programming language.
We will go over:

- Import your data into R
- Clean (tidy) data
- Transform Data
- Explore the data using visualization tools
- Run basic statistical analysis
- Communicate results

#### Basic Arithmetic

Let's try to perform basic arithmetic operations, i.e:
```{r}
4 + 7
```

We can also use brackets and basic math functions:
```{r}
3.5 + (7 - 9)/5.8 - log(1.2)
```

#### Assigning values to a variable

R uses an arrow sign to assign a value to a variable:
```{r}
x <- 2.71
```


#### Data types in R

Basic Data types in R:

- Numeric 
- Integer
- Logical
- Character

Numeric is the default data type for numbers in R:

```{r}
x <- 5.24
class(x)
typeof(x)
```

Numeric values are stored with *double* precision.

To store value as an integer, we can add *L* symbol at the end:

```{r}
i <- 3L
class(i)
typeof(i)
```

The logical data type stores boolean values:

```{r}
ans <- TRUE
class(ans)
```

The character data type variables store single characters or strings. Double or single commas can be used to define a character value:

```{r}
my.name <- "Katia Bulekova"
```
