---
title: "Research Computing Bootcamp"
author: "Boston University"
output:
  html_document:
    df_print: paged
---

<style type="text/css">

body{ /* Normal  */
  font-size: 12px;
}
td {  /* Table  */
  font-size: 8px;
}
code.r{ /* Code block */
    font-size: 12px;
    background-color: #FAEAEA; 
}
pre { /* Code block - determines code spacing between lines */
    font-size: 12px;
     
}
</style>

<br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source="bg-success")
```


## R Track: Part 1

This Bootcamp will teach you how to analyze a dataset using R programming language.
We will go over:

- Import your data into R
- Clean (tidy) data
- Transform Data
- Explore the data using visualization tools
- Run basic statistical analysis
- Communicate results

### Basic Arithmetic

Let's try to perform basic arithmetic operations, i.e:
```{r}
4 + 7
```

We can also use brackets and basic math functions:
```{r}
3.5 + (7 - 9)/5.8 - log(1.2)
```

<br/><br/>

### Assigning values to a variable

R uses an arrow sign to assign a value to a variable:
```{r}
x <- 2.71
```

We can read this line as *2.71 is assigned to x*. In many cases you can use equal sign `=` as well, but in R it is a good practice to use arrow `<-` for assignments and equal sign `=` to specify the values of arguments in functions.
<br/>


#### *Note:*

Object names in R can use letters, digits, dots, and underscores. Spaces and other special characters cannot be used to name R objects.
<br/>

```{r, eval=FALSE}

# Assign a value to a variable
x <- 2.71

# Print out the value of a variable in the console window:
print(x)

# Another way to print out the value of a variable:
x

# Assign a value to a variable and prints its value on the console
( x <- 2.71 )
```


#### **Exercise**
Assign numeric values to variables `a` and `b` and then calculate `a + b`, `a * b`, `a - b`, and `a /b `.

<br/><br/>



### Data types in R

Objects in R can have different types.

Basic Data types in R:

- **Numeric** (real numbers)
- **Integer** (whole numbers)
- **Logical** (TRUE/FALSE)
- **Character** (strings)
- **Complex** 

Numeric is the default data type for numbers in R:

```{r}
x <- 5.24
class(x)
typeof(x)
```

Numeric values are stored with *double* precision.

To store value as an integer, we can add *L* symbol at the end:

```{r}
i <- 3L
class(i)
typeof(i)
```

The logical data type stores boolean values:

```{r}
ans <- TRUE
class(ans)
```

The character data type variables store single characters or strings. Double or single commas can be used to define a character value:

```{r}
my.name <- "Katia Bulekova"
```

<br/><br/>



### R Functions

As many other languages, R has built-in functions, i.e `sqrt()`, `log()`, `round()`, etc. Many other functions are defined in R packages that you can install (we will do it a little later.)

Functions can have zero, one, or more arguments. Some arguments have *default* values:
```{r}
# List function arguments
args(round)
```
 The argument *digits* in the `round()` function has a default value 0. So if we call this function and do not use the argument *digits*, the value of `x` will be rounded to the nearest whole number. If we want to round `x` to one decimal place, we need to set *digits* to one:
```{r, eval = FALSE}
# Round to one decimal place
round(x, digits=1)
```
```{r}
# or simply:
round(x, 1)
```
 
#### *Note:*
It is  recommended to specify the argument names as it makes your code easier to understand and share with others. When you name arguments, their order does not matter!
<br/><br/>

### R Help

R comes with comprehensive help system.

Use one question mark in front of the function name if you know function's name.

Use two question marks, if you do not know function name and you search for a topic. If your search includes more than one word, use quotes:

```{r, eval = FALSE}
# Search for a help topic for round() function:
?round

# List all R help topics that include "standard deviation" phrase:
??"standard deviation"
```


#### **Exercise**
Find help topics for *working directory* phrase. From the resulting list select the function that comes from *base* package. Read the description of the function and execute it in your console window. Do not forget parenthesis. 