---
title: "Research Computing Bootcamp"
author: "Boston University"
output:
  html_document:
    df_print: paged
---

<style type="text/css">

body{ /* Normal  */
  font-size: 12px;
}
td {  /* Table  */
  font-size: 10px;
}
code.r{ /* Code block */
    font-size: 12px;
    background-color: #FAEAEA; 
}
pre { /* Code block - determines code spacing between lines */
    font-size: 12px;
     
}
</style>

<br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source="bg-success")
```


## R Track: Part 1 - Introduction to R

This Bootcamp will teach you how to analyze a dataset using R programming language.
We will go over:

- Import your data into R
- Clean (tidy) data
- Transform Data
- Explore the data using visualization tools
- Run basic statistical analysis
- Communicate results

### Basic Arithmetic

Let's try to perform basic arithmetic operations, i.e:
```{r}
4 + 7
```

We can also use brackets and basic math functions:
```{r}
3.5 + (7 - 9)/5.8 - log(1.2)
```

<br/><br/>

### Assigning values to a variable

R uses an arrow sign to assign a value to a variable:
```{r}
x <- 2.71
```

We can read this line as *2.71 is assigned to x*. In many cases you can use equal sign `=` as well, but in R it is a good practice to use arrow `<-` for assignments and equal sign `=` to specify the values of arguments in functions.
<br/>


#### *Note:*

Object names in R can use letters, digits, dots, and underscores. Spaces and other special characters cannot be used to name R objects.
<br/>

```{r, eval=FALSE}

# Assign a value to a variable
x <- 2.71

# Print out the value of a variable in the console window:
print(x)

# Another way to print out the value of a variable:
x

# Assign a value to a variable and prints its value on the console
( x <- 2.71 )
```


#### **Exercise**
Assign numeric values to variables `a` and `b` and then calculate `a + b`, `a * b`, `a - b`, and `a /b `.

<br/><br/>



### Data types in R

Objects in R can have different types.

Basic Data types in R:

- **Numeric** (real numbers)
- **Integer** (whole numbers)
- **Logical** (TRUE/FALSE)
- **Character** (strings)
- **Complex** 

Numeric is the default data type for numbers in R:

```{r}
x <- 5.24
class(x)
typeof(x)
```

Numeric values are stored with *double* precision.

To store value as an integer, we can add *L* symbol at the end:

```{r}
i <- 3L
class(i)
typeof(i)
```

The logical data type stores boolean values:

```{r}
ans <- TRUE
class(ans)
```

The character data type variables store single characters or strings. Double or single commas can be used to define a character value:

```{r}
my.name <- "Katia Bulekova"
```

<br/><br/>



### R Functions

As many other languages, R has built-in functions, i.e `sqrt()`, `log()`, `round()`, etc. Many other functions are defined in R packages that you can install (we will do it a little later.)

Functions can have zero, one, or more arguments. Some arguments have *default* values:
```{r}
# List function arguments
args(round)
```
 The argument *digits* in the `round()` function has a default value 0. So if we call this function and do not use the argument *digits*, the value of `x` will be rounded to the nearest whole number. If we want to round `x` to one decimal place, we need to set *digits* to one:
```{r, eval = FALSE}
# Round to one decimal place
round(x, digits=1)
```
```{r}
# or simply:
round(x, 1)
```
 
#### *Note:*
It is  recommended to specify the argument names as it makes your code easier to understand and share with others. When you name arguments, their order does not matter!
<br/><br/>

### R Help

R comes with comprehensive help system.

Use one question mark in front of the function name if you know function's name.

Use two question marks, if you do not know function name and you search for a topic. If your search includes more than one word, use quotes:

```{r, eval = FALSE}
# Search for a help topic for round() function:
?round

# List all R help topics that include "standard deviation" phrase:
??"standard deviation"
```


#### **Exercise**
Find help topics for *working directory* phrase. From the resulting list select the function that comes from *base* package. Read the description of the function and execute it in your console window. Do not forget parenthesis. 


<br/><br/>

### R Data Structures

R has five common data structures:<br/>

Homogeneous:

- vector
- matrix/array

Heterogeneous

- data frame
- list

The following functions help to explore data structures:<br>
`typeof()`, `length()`, `class()`, `str()`, `summary()`
<br/><br/>

### Vectors
A vector is a collection of elements of the same type (mode). Most commonly vectors are *character*, *logical*, *integer*, or *numeric*. We can use function `c()` to define a vector:

```{r}
# numeric vector
x_num <- c(5.6, 3.2, 7.8, 9.1)

# character vector
x_char <- c("copper", "iron", "silver", "lead", "gold" )

# logical vector
x_log <- c(TRUE, FALSE, FALSE, TRUE)

```

<br/>

#### **Exercise**
Define numeric, character, and logical vectors. Use 
`typeof()`, `length()`, `class()`, `str()`, `summary()` functions to explore these vectors. Compare the outputs of `summary()` function for numeric, character and logical vectors.
<br/><br/>

### Appending elements to a vector

We can add more elements to a vector using the same `c()` function:
```{r}
x_num <- c(-1.7, x_num, -2.4)
x_num
```
<br/><br/>

### Vectors from a Sequence
Using `:` operator we can create a sequence of values (an integer vector):
```{r}
( seq1 <- 1:50 )
( seq2 <- 25:4 )
```
**Note:** Both starting and ending values are included in the resulting vector.
<br/>

#### **Exercise**
Search for a help topic for function `mean()` ( `?mean` ). Using this function calculate an average value of any numeric vector you defined so far.

Try other functions.

For numeric vectors:<br/>
`max()`, `min()`, `median()`, `sd()`, `sum()`

For character vectors:<br/>
`nchar()`, `strtrim()`, `substring()`

For logical vectors:<br/>
`sum()`, `which()`
<br/><br/>

### Operations between vectors
We can perform operations on one, two or more vectors:

```{r}
# define a vector with distances in inches
heights <- c(65, 69, 60, 71, 66, 68, 64)

# convert to centimeters
heights * 2.54
```




```{r}
# define two vectors (heights in inches and weight in pounds)
heights <- c(65, 69, 60, 71, 66, 68, 64)
weights <- c(130, 142, 125, 180, 179, 154, 146)

# Calculate BMI
( bmi <- (weights/heights^2) * 703 )
```


<br/><br/>

### Vector slicing
R uses square brackets `[]` to slice/subset vectors:
```{r}
# extract second value from a vector
x_num[2]

# get all elements starting from second and ending with fourth
x_num[2:4]

#get first and fourth element
x_num[ c(1,4) ]

```

<br/><br/>

### Missing values

Missing values in R are represented with the `NA` string. It can be used for all types of vectors:

```{r}
# numeric vector
x_num <- c(5.6, 3.2, NA, 7.8, 9.1, NA)

# character vector
x_char <- c("copper", NA, "iron", "silver", "lead", NA, "gold" )

# logical vector
x_log <- c(TRUE, NA, FALSE, FALSE, NA, TRUE)

```

<br/><br/>
Functions `is.na()` and `anyNA()` help to identify if there are any missing value and if so, which one:
```{r}
# Check if there are any missing values
anyNA(x_num)

# Check each element in a vector to see which elements are missing'
is.na(x_num)

# Identify indecies of the missing elements
which( is.na(x_num) )

# Check how many missing elements
sum( is.na(x_num) )
```

<br/><br/>

Almost any computation involving a missing value will return a missing value. 
```{r}
# Calculate mean value of vector x_num
mean(x_num)
```

<br/><br/>
Many R functions have a special argument, that allows to remove all missing values:
```{r}
# Calculate mean value of vector x_num ignoring missing values
mean(x, na.rm=TRUE)
```

<br/>

#### **Exercise**
Apply `summary()` function to any numeric vector you defined that has missing values. Examine the output. 
<br/><br/>

### Factors

Factors in R are used to store categorical data. When R vectors are converted to factors, each unique value is assigned an integer value (label). By default these integer labels are assigned alphabetically:
```{r}
# define a vector:
ans <- c("yes", "no", "no", "yes", "yes", "yes")
summary(ans)

# convert vector to factor:
ans <- factor(ans)
summary(ans)
```

<br/>
Sometimes, categorical data are coded as integer values (for example 1-for female, 2-for male). It makes sense to convert them to factors and assign appropriate labels:
```{r}
# gender defined with integer values
gender <- c(1, 2, 2, 1, 1, 1, 2, 2, 1, 2, 2, 2)
summary (gender)

# convert vector to factor and assign appropriate labels
gender <- factor( gender, labels = c("female", "male"))
summary(gender)
```


### Reading Data

R can read data written in many different formats. Today we will read basic CSV file.
You can see the following link to see examples of reading data saved in different formats:

```{r}
# Read CSV file


# convert vector to factor and assign appropriate labels
gender <- factor( gender, labels = c("female", "male"))
summary(gender)
```
